---
title: "Babies vs babies"
author: "Andrew MacDonald"
description: |
  Simulations of parasitized baby birds, with varying degrees of realism.
date: 06 Oct 2023
draft: true
editor: source
categories: [UdeS, stan]
---

```{r setup, eval=TRUE, message=FALSE, warning=FALSE}
library(targets)
library(ggplot2)
library(tidyverse)
library(tidybayes)
```

Calculating GDDs but its baby birds getting eaten alive by maggots

```{r}
ois_list <- rerun(100, {
  taille_couvee <- sample(3:8, 1)
  pr_surv_ois <- 0.98
  temps_nid <- sample(16:22, 1)
  n_ois <- numeric(temps_nid)
  n_ois[1] <- taille_couvee
  for(t in 2:temps_nid){
    n_ois[t] <- rbinom(1, n_ois[t-1], prob = pr_surv_ois)
  }
  n_ois
})
```

assumption: nestling survival is independent of parasitism (its an hypothesis later) -- can exaggerate stresses.


```{r}
ois_df <- ois_list |> 
  map_df(~ tibble(nois = ., jour = 1:length(.)), .id = "nid_id")

ois_df |> 
  ggplot(aes(x = jour, y = nois, group = nid_id)) + geom_line() + facet_wrap(~nid_id)
```


parasites

```{r}
ois_df |> filter(nid_id ==1) |> as.data.frame()
```


What is the relationship between number of chicks and prob of parasites??

```{r}
x_ois <- 0:8
avg_logit <- -4
slope_logit <- .3
prob_parasite <- exp(avg_logit + slope_logit * x_ois)/(1 + exp(avg_logit + slope_logit*x_ois))
plot(x_ois, prob_parasite, type = "l", ylim = c(0, 1))
```

chance of getting a parasite every day

```{r}
ois_para_df <- ois_df |> 
  mutate(probab_para = exp(avg_logit + slope_logit*nois)/(1 + exp(avg_logit + slope_logit*nois)),
         actual_para = rbinom(length(probab_para), 1, prob = probab_para)
         )
```

assuming that parasitized nests can get MORE parasites

parasites need to arrive like 4 days before fledging to survive and get counted

```{r}
parasite_results <- ois_para_df |> 
  mutate(parasites_safe = (jour <= (max(jour) - 4)) & nois>0) |> 
  group_by(nid_id) |> 
  summarise(chick_avail = sum(nois),
            para_pres = sum(actual_para * parasites_safe)>0)
```

```{r}
parasite_results |> 
  ggplot(aes(x = chick_avail, y = as.numeric(para_pres))) + geom_count()  + 
  stat_smooth(method = glm, method.args = list(family = "binomial"))
```

next steps, 
* make a shiny app?
things to change:

* variation over the time over different ages -- older chicks die less, and are less attractive to parasites

particularly interesting is the relationship btween the attractiveness of chicks 
is ther ea way to make a logistic curve or another curve pounded by 0 and 1, that starts at 0 at a known amount? I think the modified logistic was kind of like that

I need ot read more actualy papers, I'm getting inward looking and dumb

let me start with the papers of the scientists that I am wlorking with -- i'll read all of Matt's papers, and then all of the papers - or the most recent papers - of the scientists in McGill and  udes. 

But this is only a part of the story.
we've shown that at least some of the increasting chick-days 

hm could you also use a type 2 predator response? something with a diminshing return for the number of eggs in a nest? I imagine that mahy of these insect species area actually tracking the number of indiviaul, to a total above 8, since they propbably also atack birds with bigger clutch sizes

ohh what about biomass,, are they flies atually attracted to THAT -- and what is the relationhsip between biomass and density in chicks? do chicks die as the nest gets crowded? probably not, but I nw that happens sometimes. 

in general, do people relate growth, onteogeny, self thinning to colonization rates? 

and how do we think about indicence when the area gets bigger, anyway? 


