---
title: "Phylogeny"
author: "Andrew MacDonald"
description: |
  it is that (evolutionary) time.
date: 27 Sept 2024
editor: source
categories: [UdeS, stan]
draft: false
editor_options: 
  chunk_output_type: console
bibliography: references.bib
---

```{r setup, eval=TRUE, message=FALSE, warning=FALSE}
library(targets)
library(ggplot2)
library(tidyverse)
library(tidybayes)
```

I've always wanted to understand how to do phylogenetic regressions, and thanks to my colleauge Alex Fuster I finally had the opportunity to sit down and work on them. The literature on the topic is *confusing*, large, and not always clear about what model is being fit. I relied heavily on two resources:

-   one book, a free online resource from Tony Ives [@ives]
-   a preprint, an excellent review of the use and misuse of Pagel's Lambda by @pearse

## Why a phylogenetic regression?

Suppose you have two traits, measured across many different species – say, social group size (Trait X) and brain size (Trait Y). You want to test the hypothesis that bigger social groups mean a bigger brain. However there's a catch: some of the species are closely related, and others are not. Its entirely possible that any apparent correlation between Trait X and Trait Y comes from random chance: both traits change randomly along evolutionary time. That means that distantly related species have more time to become different to each other, and close relatives have less "time apart" and are therefore less likely to be different in their two traits.

Because every kind of cross-species comparison involves a group of species with a phylogenetic structure, "controlling for phylogeny" has become very common in these kinds of studies. Also, because we are usually missing traits for at least some species in our studies, people often use phylogeny as a guide for guessing what trait values are present in the animals that we haven't measured.

## Recipe for phylogeny

I love the large and flexible toolbox of Bayesian methods because it can be adapted to fit such a huge array of models – virtually all the models that ecologists want to fit! However, there's a catch: to fit a model using Stan (or something similar) you have to know *exactly* what model you're fitting. However, because these regressions are usually fit using custom software, it can be a challenge to dig to find the exact equations being fit!

Using the two resources mentioned above, I was able to write down (I hope!) the equation for the model like this:

$$
\begin{align}
y_i &\sim \text{Normal}(\bar{y} + \beta_1 x_i + a_{s[i]}, \sigma_y) \\
a_{s} &\sim \text{MVNormal}(0, \Sigma_a)\\
\Sigma_a &= \begin{bmatrix}
\sigma_a^2 & \lambda_a \cdot \sigma_{12} & \cdots & \lambda_a \cdot \sigma_{1,s} \\
\lambda_a \cdot \sigma_{21} & \sigma_a^2 & \cdots & \lambda_a \cdot \sigma_{2,s} \\
\vdots & \vdots & \ddots & \vdots \\
\lambda_a \cdot \sigma_{s,1} & \lambda_a \cdot \sigma_{s,2} & \cdots & \sigma_a^2
\end{bmatrix} \\

x_i &\sim \text{Normal}(\bar{x} + b_{s[i]}, \sigma_x) \\
b_{s} &\sim \text{MVNormal}(0, \Sigma_b)\\
\Sigma_b &= \begin{bmatrix}
\sigma_b^2 & \lambda_b \cdot \sigma_{12} & \cdots & \lambda_b \cdot \sigma_{1,s} \\
\lambda_b \cdot \sigma_{21} & \sigma_b^2 & \cdots & \lambda_b \cdot \sigma_{2,s} \\
\vdots & \vdots & \ddots & \vdots \\
\lambda_b \cdot \sigma_{s,1} & \lambda_b \cdot \sigma_{s,2} & \cdots & \sigma_b^2
\end{bmatrix}



\end{align}
$$ {#eq-phyloreg}


You can see that there are two big variance-covariance matrices here, for the effects of phylogeny on $y$ and $x$. 
These covariance matrices have three ingredients that are all put together: 

1. _the base_ How far apart are species on the phylogeny? Many ecologists work with trees where all the tips end at the present day -- so all species have the same amount of time going back to their last common ancestor. For trees like this, the diagonal is 1 (i.e., 100% of the evolutionary time). The off-diagonals are the proportion of this total time which is shared between species.
2. _the flavour_ This is a model of species averages. If there were no effect of phylogeny at all, we would still expect species to be a little different. But how different are species from each other? That is controlled by a standard deviation, $\sigma$, which we multiply the whole matrix by to scale it. 
3. _the secret sauce_ The off-diagnal elements of $\Sigma$ are multiplied by **another** number between 0 and 1: this is "Pagel's Lambda". It acts like a tuning knob, adjusting the amount of phylogenetic flavour that makes it into the model. When $\lambda$ is 1, we have the maximum amount of covariance coming from the phylogeny. When $\lambda$ is 0, we are back to an identity matrix and the species are independent. 

There's another way to write this equation that makes these three parts more clear to see. First we have to make $V_{phy}$, which is the _phylogenetic variance-covariance matrix_. This has variances and covariances for each species on our tree. For example, for 3 species the phylogenetic variance covariance matrix is:

$$
V_{phy} = \begin{bmatrix}
\sigma_1^2 & \sigma_{12} & \sigma_{1,3} \\
\sigma_{2,1} & \sigma_2^2 & \sigma_{2,3} \\
\sigma_{3,1} & \sigma_{3,2} & \sigma_3^2
\end{bmatrix}
$$
The covariances are equal to the proportion of the tree that is shared between two species. The diagonal is the amount of time between the tree's start and each species. This is just a fancy way to say that, for a tree where all the tips end at the present day, the diagonal is 1 and the off-diagonal is between 0 and 1.

Then, we can write the expression for $\Sigma$ like this:

$$
\Sigma = \sigma^2 \lambda V_{phy} + \sigma^2 (1 - \lambda) \mathbf{I}
$$
This is equation 4 in @pearse. 


I can rewrite @eq-phyloreg in this style:

$$ 
\begin{align}

y_i &\sim \text{Normal}(\bar{y} + \beta_1 x_i + a_{s[i]}, \sigma_y) \\
a_{s} &\sim \text{MVNormal}(0, \Sigma_a)\\
\Sigma_a &= \sigma_a^2 \lambda_a V_{phy} + \sigma_a^2 (1 - \lambda_a) \mathbf{I} \\
x_i &\sim \text{Normal}(\bar{x} + b_{s[i]}, \sigma_x) \\
b_{s} &\sim \text{MVNormal}(0, \Sigma_b)\\
\Sigma_b &= \sigma_b^2 \lambda_b V_{phy} + \sigma_b^2 (1 - \lambda_b) \mathbf{I} \\
\end{align}

$$
You can see I'm using two different trait variances and two different $\lambda$s, because the two traits might have a different amount of variance among species, and within that might have a different amount of phylogenetic signal.


## Data simulation

Here is simulation code from @ives , which generates a dataset where there is a signal for phylogeny and also a relationship between two traits of interest. I'll use this code to generate a dataset and then estimate the known parameters with a Stan model:


```{r}
## simulate data
set.seed(1618)
n <- 20
b0 <- 0
b1 <- -.1
lam.x <- .98
lam.e <- .8
sigma_y <- .6
sigma_x <- 1

phy <- ape::compute.brlen(
  ape::rtree(n=n),
  method = "Grafen",
  power = 1)

plot(phy)

phy.x <- phylolm::transf.branch.lengths(
  phy=phy, model="lambda",
  parameters=list(lambda = lam.x))$tree

phy.e <- phylolm::transf.branch.lengths(
  phy=phy, model="lambda",
  parameters=list(lambda = lam.e))$tree

x <- ape::rTraitCont(phy.x, model = "BM", sigma = sigma_x)
e <- ape::rTraitCont(phy.e, model = "BM", sigma = sigma_y)
x <- x[match(names(e), names(x))]
Y <- b0 + b1 * x + e
Y <- array(Y)
rownames(Y) <- phy$tip.label

plot(x, Y)

```

```{r}
#| class-output: stan
phylo <- cmdstanr::cmdstan_model(here::here("posts/2024-09-27-phylo/phylo.stan"))

phylo
```

```{r, warning=FALSE, message=FALSE}
phylo_sample <- phylo$sample(data = list(
  n = n,
  s = n,
  x = x,
  y = Y,
  phyvcv = ape::vcv(phy)
),parallel_chains = 4, refresh = 1000)

phylo_sample
```





## Tips from the forum

```{r}
#| class-output: stan
phylo_forum <- cmdstanr::cmdstan_model(here::here("posts/2024-09-27-phylo/phylo_forum.stan"))

phylo_forum
```

```{r}
phylo_forum_sample <- phylo_forum$sample(data = list(
  n = n,
  s = n,
  x = x,
  y = Y,
  phyvcv = ape::vcv(phy)
),parallel_chains = 4, refresh = 1000)

phylo_forum_sample
```

and an even simpler strategy, replacing the lambda parameter on the logit scale with a beta:

```{r}
#| class-output: stan
phylo_beta <- cmdstanr::cmdstan_model(here::here("posts/2024-09-27-phylo/phylo_beta.stan"))

phylo_beta
```

```{r}
phylo_beta_sample <- phylo_beta$sample(data = list(
  n = n,
  s = n,
  x = x,
  y = Y,
  phyvcv = ape::vcv(phy)
),parallel_chains = 4, refresh = 1000)

phylo_beta_sample
```

## Repeated sampling of these traits

The simulation above is giving species *means*. However in our study we have more than one measurement per species. Measurements of "Trait X" and "Trait Y" are measured on different individuals.

```{r}
library(tidyverse)
require(ape)

set.seed(1618)

# set true parameter values
n <- 20
b0_y <- .5
b1 <- .5
lam.x <- .98
lam.e <- .5
sigma_x <- 1
sigma_y <- .3

# simulate phylogeny
phy <- ape::compute.brlen(
  ape::rtree(n=n),
  method = "Grafen",
  power = 1)

plot(phy)


# get names from this matrix! needs to line up perfectly
phyvcv <- ape::vcv(phy)

distmat_names <- dimnames(phyvcv)[[1]]

# observations per species
n_obs <- 15


phy.x <- phylolm::transf.branch.lengths(
  phy=phy, model="lambda",
  parameters=list(lambda = lam.x))$tree

phy.e <- phylolm::transf.branch.lengths(
  phy=phy, model="lambda",
  parameters=list(lambda = lam.e))$tree

x <- ape::rTraitCont(phy.x, model = "BM", sigma = sigma_x)
e <- ape::rTraitCont(phy.e, model = "BM", sigma = sigma_y)
x <- x[match(names(e), names(x))]

## calculate Y
Y <- b0_y + b1 * x + e

# Y <- array(Y)
names(Y) <- phy$tip.label

plot(x, Y)


obs_xy_df <- tibble(x, Y, sp_name = names(x)) |> 
  mutate(
    sp_id = as.numeric(
      factor(sp_name, 
             levels = distmat_names))) |> 
  rowwise() |> 
  mutate(obs_x = list(
    rnorm(n_obs, mean = x, sd = .3)),
    obs_y = list(rnorm(n_obs, mean = Y, sd = .3)))


x_obs_df <- obs_xy_df |> 
  select(sp_id, obs_x) |> unnest(obs_x)

y_obs_df <- obs_xy_df |> 
  select(sp_id, obs_y) |> unnest(obs_y)
```

fit a model that is ready for replication per species:

```{r}
#| class-output: stan
phylo_obs <- cmdstanr::cmdstan_model(here::here("posts/2024-09-27-phylo/phylo_obs.stan"))

phylo_obs
```

```{r}
phylo_obs_sample <- phylo_obs$sample(data = list(
  s = n,
  # trait x
  n_x = nrow(x_obs_df),
  x_obs = x_obs_df$obs_x,
  sp_id_x = x_obs_df$sp_id,
  # trait y
  n_y = nrow(y_obs_df),
  y_obs = y_obs_df$obs_y,
  sp_id_y = y_obs_df$sp_id,
  # phylogeny
  phyvcv = phyvcv
),parallel_chains = 4, refresh = 1000)

phylo_obs_sample$summary(variables = c(
  "b0_y", "b1", "sigma_x", "sigma_y", "lambda_x", "lambda_y", "sigma_x_obs", "sigma_y_obs"
))
```

I guess I thought it would be better to sample `lambda` on the normal scale as well, but is that misguided? I guess using the unconstrained scale also sets us up for hierarchical effects later.

In the real-world work that this is focussed on, the traits X and Y are going to be observed on different individuals. Sample size will be different across species with some species missing entirely from one or the other (or both) X and Y.
