---
title: "Validating a model of selection on plasticity"
author: "Andrew MacDonald"
description: "Plus ça change, plus c'est la change qui change"
date: 2023-02-02
categories: [UdeS, stan]
execute:
  eval: false
---

```{r setup, eval=TRUE, message=FALSE, warning=FALSE}
library(cmdstanr)
library(tidyverse)
library(tidybayes)
```


Let's begin just with a simulation of three interrelated traits:

* when a bird arrives, which determines
* how many eggs they lay, out of which there is 
* some number of surviving offspring

```{r}
## how many birds
nbirds <- 57
# simulate arrival dates -- two weeks before and after whatever the average is
darrive <- runif(nbirds, min = -14, max = 14) |> round()

## simulate clutch sizes -- decrease by 4% each day
avg_clutch <- 4.5
effect_per_day <- .97

clutch <- rpois(nbirds, exp(log(avg_clutch) + log(effect_per_day)*darrive))

plot(darrive, clutch)
```

As an aside, this would be 0-truncated, since birds who don't lay eggs don't get observed at all.

```{r}
# simulate hatching success
success <- rbinom(nbirds, size = clutch, prob = .86)

plot(darrive, success)
```

what's important to see here is that there is a negative correlation, even though the arrival date has no direct effect on the outcome

```{r}
summary(glm(success ~ darrive, family = "poisson"))
```


But, if we use a binomial model that knows about the number of possible successful chicks, then we see what we expect: 

```{r}
bin_mod <- (glm(cbind(success, clutch - success) ~ 1, family = binomial(link = "logit")))

plogis(coef(bin_mod))
```

Which matches the simulation above. 

if we put `darrive` in the model, the effect should be very close to 0 with overlap.

If we imagine that the laying date effects the survival p, then we should see an effect close to 0

```{r}
summary(glm(
  cbind(success, clutch - success) ~ 1 + darrive,
            family = binomial(link = "logit")))
```

One Stan model can model all of these at the same time

```{r}
#| class-output: stan

one_indiv <- cmdstanr::cmdstan_model(
  here::here("posts/2023-02-02-selection-on-plasticity/one_indiv.stan"))

one_indiv
```


```{r}
one_indiv_post <- one_indiv$sample(
  data = list(nbirds = nbirds,
              clutch = clutch, 
              success = success, 
              darrive = darrive))

one_indiv_post
```

reasonably close to true values:

```{r}
plogis(1.44)
log(avg_clutch)
log(effect_per_day)
```


```{r}

simulate_some_birds <- function(nbirds = 57, 
         log_b_date = log(.97),
         log_avgclutch = log(4.5),
         logit_psuccess = qlogis(.84)){
  
  # simulate arrival dates -- two weeks before and after whatever the average is
  darrive <- runif(nbirds, min = -14, max = 14) |> round()
  
  ## simulate clutch sizes -- decrease  each day
  clutch <- rpois(nbirds, exp(log_avgclutch + log_b_date*darrive))
  
  ## simulate success
  success <- rbinom(nbirds, size = clutch, prob = plogis(logit_psuccess))
  
  return(list(
    data_list = list(
      nbirds = nbirds,
      darrive = darrive, 
      clutch = clutch, 
      success = success 
    ),
    true_values = tribble(
      ~variable, ~true_value,
      "log_b_date", log_b_date,
      "log_avgclutch", log_avgclutch,
      "logit_psuccess", logit_psuccess
    )
  ))
}


data_for_simulation <- simulate_some_birds()

one_indiv_post <- one_indiv$sample(data = data_for_simulation$data_list,
                                   refresh = 0)

comparison <- one_indiv_post |> 
  # tidybayes::gather_rvars(logit_psuccess, log_avgclutch, log_b_date) |> 
  tidybayes::tidy_draws() |> tidybayes::gather_variables() |> 
  right_join(data_for_simulation$true_values, by = c(".variable" = "variable"))


comparison |> 
  ggplot(aes(y = .variable, x = .value)) + 
  stat_halfeye() + 
  geom_point(aes(y = variable, x = true_value),
             col = "orange",
             pch = "|",
             size = 10, data = data_for_simulation$true_values)
```

## No 0 birds

This system is a little challenging, since we never observe 0 eggs per bird -- if a bird cannot lay eggs (e.g. it does not find a nest spot) then it goes uncounted

To simulate this, I'll drop the 0 clutches before doing the rest of the simulations. This means that sample size will be less than or equal to the "nbirds" argument.

```{r}

simulate_some_birds_nonzero <- function(nbirds = 57, 
         log_b_date = log(.97),
         log_avgclutch = log(4.5),
         logit_psuccess = qlogis(.84)){
  
  # simulate arrival dates -- two weeks before and after whatever the average is
  darrive <- runif(nbirds, min = -14, max = 14) |> round()
  
  ## simulate clutch sizes -- decrease  each day
  clutch <- rpois(nbirds, exp(log_avgclutch + log_b_date*darrive))
  # drop 0 nests
  nonzero_clutch <- which(clutch > 0)
  
  ## simulate success
  success <- rbinom(nbirds, size = clutch, prob = plogis(logit_psuccess))
  
  return(list(
    data_list = list(
      nbirds = length(nonzero_clutch),
      darrive = darrive[nonzero_clutch], 
      clutch  =  clutch[nonzero_clutch], 
      success = success[nonzero_clutch] 
    ),
    true_values = tribble(
      ~variable, ~true_value,
      "log_b_date", log_b_date,
      "log_avgclutch", log_avgclutch,
      "logit_psuccess", logit_psuccess
    )
  ))
}

set.seed(1234)
some_nonzeros <- simulate_some_birds_nonzero(nbirds = 200)

```

a plot to confirm that it works:

```{r}
some_nonzeros$data_list |> 
  as.data.frame() |> 
  ggplot(aes(x = darrive, y = clutch)) + geom_point()
```

And fit the posterior

```{r}
plot_posterior_true <- function(simdata, stanmodel){
  model_post <- stanmodel$sample(data = simdata$data_list,
                                     refresh = 0, parallel_chains = 4)
  
  comparison <- model_post |> 
    tidybayes::tidy_draws() |> 
    tidybayes::gather_variables() |> 
    right_join(simdata$true_values, by = c(".variable" = "variable"))
  
  
  comparison |> 
    ggplot(aes(y = .variable, x = .value)) + 
    stat_halfeye() + 
    geom_point(
      aes(y = variable, x = true_value),
      col = "orange",
      pch = "|",
      size = 10, data = simdata$true_values)
}


plot_posterior_true(some_nonzeros, one_indiv)

```

There's already some bias happening! let's try what happens when the average is lower (and gives more 0s)

```{r}
set.seed(420)
simulate_some_birds_nonzero(log_avgclutch = log(2.4),
                            log_b_date = log(.7),
                            nbirds = 300) |> 
  plot_posterior_true(one_indiv)
```

some preliminary repetitions show that it usually misses either the average or the hatching success, frequently both. 

```{r}
#| class-output: stan

one_indiv_noZero <- cmdstanr::cmdstan_model(
  here::here("posts/2023-02-02-selection-on-plasticity/one_indiv_noZero.stan"))

one_indiv_noZero
```

```{r}

set.seed(420)
simulate_some_birds_nonzero(log_avgclutch = log(4.4),
                            log_b_date = log(.9),
                            nbirds = 300) |> 
  plot_posterior_true(one_indiv_noZero)
```

### Truncating using a different syntax

The manual [uses a different syntax](https://mc-stan.org/docs/stan-users-guide/truncated-data.html). 
To write the equation above this way, I found I needed to make a few changes:

* `poisson_log` has to be replaced with `poisson` and the `exp()` link function

... that was actually the only change. 
It runs at the same speed as the previous way of writing it, and gets the same answer:

```{r}
#| class-output: stan

one_indiv_zerotrunc <- cmdstanr::cmdstan_model(
  here::here("posts/2023-02-02-selection-on-plasticity/one_indiv_zerotrunc.stan"))

one_indiv_zerotrunc
```

```{r}

set.seed(420)
simulate_some_birds_nonzero(log_avgclutch = log(4.4),
                            log_b_date = log(.9),
                            nbirds = 300) |> 
  plot_posterior_true(one_indiv_zerotrunc)
```

:::{.callout-note}
# Two parameterizations diverged in a yellow wood
These two ways of writing the model both work. Which to choose? 
Well, I was pleased with myself for manually normalizing the Poisson likelihood in the first model. 
However the second is clearer to the reader. In the first, it takes two lines of code -- not even necessarily next to each other. In the second, the big `T` for *T*runcation indicates clearly what is going on. Code is communication; clarity wins.
:::

### Zero inflated binomial success

Once in a while, a nest will just be completely destroyed by, say, a predator. 
This has nothing to do with anything, probably, and is just a catastrophic Act of Weasel. 
Let's imagine that some small proportion of the nests just die:

```{r}

simulate_some_birds_nonzero_zeroinflated <- function(nbirds = 57, 
         log_b_date = log(.97),
         log_avgclutch = log(4.5),
         logit_psuccess = qlogis(.84),
         logit_pfail = qlogis(.1)){
  
  # simulate arrival dates -- two weeks before and after whatever the average is
  darrive <- runif(nbirds, min = -14, max = 14) |> round()
  
  ## simulate clutch sizes -- decrease  each day
  clutch <- rpois(nbirds, exp(log_avgclutch + log_b_date*darrive))
  # drop 0 nests
  nonzero_clutch <- which(clutch > 0)
  n_laid <- length(nonzero_clutch)
  
  # simulate success -- among birds which laid at least 1 egg, there is a chance of failing completely
  success_among_nonzero <- rbinom(n_laid,
                                  size = clutch[nonzero_clutch],
                                  prob = plogis(logit_psuccess))
  failed_nests <- rbinom(n_laid, size = 1, prob = plogis(logit_pfail))

  success_zi <- success_among_nonzero * (1 - failed_nests)
  
  # success <- rbinom(nbirds, 
  #                   size = clutch,
  #                   prob = plogis(logit_psuccess))
  # failed_nests <- rbinom(nbirds, size = 1, prob = plogis(logit_pfail))
  # 
  # success_zi <- success * (1 - failed_nests)
  
  return(list(
    data_list = list(
      nbirds = n_laid,
      darrive = darrive[nonzero_clutch], 
      clutch  =  clutch[nonzero_clutch], 
      success = success_zi#[nonzero_clutch]
    ),
    true_values = tribble(
      ~variable, ~true_value,
      "log_b_date", log_b_date,
      "log_avgclutch", log_avgclutch,
      "logit_psuccess", logit_psuccess,
      "logit_pfail", logit_pfail
    )
  ))
}

```


```{r}
#| class-output: stan

one_indiv_ztrunc_zinf <- cmdstanr::cmdstan_model(
  here::here("posts/2023-02-02-selection-on-plasticity/one_indiv_ztrunc_zinf.stan"))

one_indiv_ztrunc_zinf
```

```{r}

set.seed(477)
some_zi_birds <- simulate_some_birds_nonzero_zeroinflated(log_avgclutch = log(4.4),
                            log_b_date = log(.9),
                            nbirds = 300) 

some_zi_birds|> 
  plot_posterior_true(one_indiv_ztrunc_zinf)
```


I initially failed to recover the parameter for `logit_pfail`. Here I some things I learned: 

* Simulating zero-inflated numbers can be tricky! I flipped back and forth between simulating 0-inflation for all nests, and simulating for only those with at least 1 egg. In retrospect, it is clear that these are equivalent. There are two independent things here: the probability of a clutch having 0 eggs and the probability of a 0 coming from the zero-inflated binomial. 
* this zero-inflated model is quite _sensitive to the prior_. That's because there are not very many zeros being inflated: in my simulation at least, failed nests are rare and success is naturally low. it would be pretty important to decide in advance if sudden nest failure (e.g. by predation) is rare or common
