---
title: "Including measurement uncertainty"
author: "Andrew MacDonald"
description: |
  Sometimes we have slightly-known knowns.
date: 11 Nov 2022
editor: source
categories: [UdeS, stan]
draft: true
---

```{r setup, eval=TRUE, message=FALSE, warning=FALSE}
library(targets)
library(ggplot2)
library(tidyverse)
library(tidybayes)
library(cmdstanr)
```

## Two ways of thinking about measurement error

* repeat independent measures of your predictor variable
* standard error of the x variables 

let's simulate each


## repeated measures of your predictor


Photosynthesis is often limited by phosphorous -- more P, more chlorophyll in lake water.
let's say you're measuring phosphorous in lakes. For each lake you have one measure of primary productivity (chlorophyll concentration) and 5 measures of phosphorous (from water samples). 
The P actually causes the chlorophyll -- but the _true_ average P concentration also causes the measurements we are making!

```{r}
avg_chl <- 60
chl_per_P <- 5
sd_chl <- 3
sd_P <- .7

chl_p <- tibble(Phos = seq(from = -1, to = 5, length.out = 25), 
       chl = rnorm(n = 25, mean = avg_chl + chl_per_P * Phos, sd = sd_chl),
       P_meas = map(Phos, ~ rnorm(n = 5, mean = .x, sd = sd_P)))

chl_p  |> 
  unnest(P_meas) |> 
  ggplot(aes(x = P_meas, y = chl)) + geom_point()
```


```{r}
#| class-output: stan
repeat_measures <- cmdstanr::cmdstan_model(
  here::here("posts/2023-09-01-measurement-error/repeat_measures.stan"),
  pedantic = TRUE)
repeat_measures
```

```{r}
repeat_measures_post <- repeat_measures$sample(
  data = list(
    n = 25, 
    P = chl_p$P_meas,
    chl = chl_p$chl
  ),refresh = 0L)
```

plot the line

plot the intervals for the two slope parameters

plot the true P values

## using SD and mean values

```{r}
#| class-output: stan
sd_mean_only <- cmdstanr::cmdstan_model(
  here::here("posts/2023-09-01-measurement-error/sd_mean_only.stan"), pedantic = TRUE)
sd_mean_only
```

```{r}
sd_mean_only_post <- sd_mean_only$sample(
  data = list(
    n = 25, 
    P_sds = map_dbl(chl_p$P_meas, sd),
    P_means = map_dbl(chl_p$P_meas, sd),
    chl = chl_p$chl
  ),refresh = 0L)
```


```{r}
#| label: fig-compare
#| fig-cap: "The posterior distribution for the parameters is the same both ways"
#| fig-subcap: 
#|   - "using means and SDs"
#|   - "using repeated measurements"
#| layout-nrow: 2
sd_mean_only_post$draws() |> 
  bayesplot::mcmc_areas(c("intercept", "slope")) + 
  labs(title = "from means and sd")


repeat_measures_post$draws() |> 
  bayesplot::mcmc_areas(c("intercept", "slope")) + 
  labs(title = "from repeated measures")
```


## comparing inference from the two




