---
title: "Ordered contrasts in two moments"
author: "Andrew MacDonald"
description: |
  it's giving Goldilocks
date: 11 Nov 2022
editor: source
draft: true
categories: [UdeS, stan]
---

```{r setup, eval=TRUE, message=FALSE, warning=FALSE}
library(targets)
library(ggplot2)
library(tidyverse)
library(tidybayes)
library(cmdstanr)
```

## What this post is about

I'm passionate about ordered contrasts. I think that very often ecologists have experimental treatments that are in the form of three levels:  less than normal, normal amount, more than normal amount

lets program this into Stan

```{r}
lil_fac <- letters[1:3]
lil_fac <- ordered(lil_fac)
ord_contr <- cbind(1, contrasts(lil_fac))
dimnames(ord_contr) <- NULL
# dput(ord_contr)
```


```{r}
ordinal_mean_prior <- cmdstan_model(here::here("posts/2023-08-29-ordinal-moments/ordinal_mean_prior.stan"))

ordinal_mean_prior
```


```{r}
ordinal_mean_prior_predictive <- ordinal_mean_prior$sample(data = list(n_each = 10))
```


```{r}
ordinal_mean_prior_predictive$draws() |> 
  gather_draws(obs[i], ndraws = 6) |> 
  ggplot(aes(x = rep(rep(1:3, each = 10), times = 6), y = .value)) + geom_point() + 
  facet_wrap(~ .draw)
```

Writing that whole thing in Stan was needlessly painful! here is an example with the id of the group instead (1, 2, or 3)

```{r}
#| class-output: stan
ordinal_grpid_prior <- cmdstan_model(here::here("posts/2023-08-29-ordinal-moments/ordinal_grpid_prior.stan"))

ordinal_grpid_prior
```


```{r}
ordinal_grpid_prior_predictive <- ordinal_grpid_prior$sample(data = list(n = 30, group_id = rep(1:3, each = 10)))

ordinal_grpid_prior_predictive$draws() |> 
  gather_draws(obs[i], ndraws = 6) |> 
  ggplot(aes(x = rep(rep(1:3, each = 10), times = 6), y = .value)) + geom_point() + 
  facet_wrap(~ .draw)
```

## condition on (fake) observations

its the same model as before except we move `obs` up to the data block:

```{r}
#| class-output: stan
ordinal_contrasts <- cmdstan_model(here::here("posts/2023-08-29-ordinal-moments/ordinal_contrasts.stan"))

ordinal_contrasts
```


We can try this out on simulated data to confirm that it matches what we'd like to see:

```{r}
means <- c(3,4,9)

neach <-  25
ys <- rnorm(n = neach*length(means), mean = rep(means, each = neach), sd = .3)
group_ids <- rep(1:3, each = neach)


ordinal_contrasts_post <- ordinal_contrasts$sample(data = list(n = length(ys),
                                                               group_id = group_ids,
                                                               obs = ys))


ordinal_contrasts_post$draws() |> 
  gather_rvars(yrep[i]) |> 
  ggplot(aes(x = i, ydist = .value)) + 
  stat_pointinterval() + 
  geom_jitter(aes(x = group_ids, y = ys), 
              width = .2, alpha = 0.3,
             data = data.frame(group_ids, ys), 
             inherit.aes = FALSE, col = "red")
```

## concrete example

Let's consider the growth rate of seedlings in proportion to the distance from a parent plant. 
We'd expect this to decline as you get farther away, and probably in a nonlinear way..

```{r}
true_values <-  c(5, -3, 1.1)
gr_mu <- ord_contr %*% true_values

plot(gr_mu)

```

::: .{callout-note}
See how easy that was to make!? that's why this works! ordered contrasts are easier to think about. 
:::

```{r}
growth_data <- tibble(group_id = rep(1:3, each = 30), 
                      true_mean = gr_mu[group_id],
                      distance = c("1m", "5m", "20m")[group_id],
                      obs = rnorm(length(true_mean), mean = true_mean, sd = 1.1))

growth_data |> 
  ggplot(aes(x = distance, y = obs)) + geom_point()
```


fit it with the same model

```{r}
ordinal_contrasts_post <- ordinal_contrasts$sample(data = list(n = nrow(growth_data),
                                                               group_id = growth_data$group_id,
                                                               obs = growth_data$obs))

ordinal_contrasts_post$summary(variables = "betas")
true_values
```

So we get very very close to the true values! 


## on the second moment also

would that even work

```{r}
#| class-output: stan
ordinal_contr_mu_sd <- cmdstan_model(here::here("posts/2023-08-29-ordinal-moments/ordinal_contr_mu_sd.stan"))

ordinal_contr_mu_sd
```


```{r}
ordinal_contr_mu_sd_post <- ordinal_contr_mu_sd$sample(data = list(n = nrow(growth_data),
                                                               group_id = growth_data$group_id,
                                                               obs = growth_data$obs))

ordinal_contr_mu_sd_post$summary(variables = c("mu_beta", "sigma_beta"))
true_values
```


```{r}
#| class-output: stan
ordinal_contr_mu_sd_prior <- cmdstan_model(here::here("posts/2023-08-29-ordinal-moments/ordinal_contr_mu_sd_prior.stan"))

ord_mu_sd_sample <- ordinal_contr_mu_sd_prior$sample(data = list(
  n = nrow(growth_data),
  group_id = growth_data$group_id))

some_samples <- ord_mu_sd_sample$draws(variables = "obs") |> 
  tidybayes::spread_draws(obs[i], ndraws = 12) |> 
  mutate(group = growth_data$group_id[i])

some_samples |> 
  ggplot(aes(x = group, y = obs)) + 
  geom_point() + 
  facet_wrap(~.iteration)
```   

## Mean - variance correlation

```{r}

ordinal_contr_mu_sd_matrix_prior <- cmdstan_model(here::here(
  "posts/2023-08-29-ordinal-moments/ordinal_contr_mu_sd_matrix_prior.stan"))

ord_mu_sd_sample <- ordinal_contr_mu_sd_matrix_prior$sample(data = list(
  n = nrow(growth_data),
  group_id = growth_data$group_id), chains = 1, refresh = 0)

some_samples <- ord_mu_sd_sample$draws(variables = "obs") |> 
  tidybayes::spread_draws(obs[i], ndraws = 12) |> 
  mutate(group = growth_data$group_id[i])

some_samples |> 
  ggplot(aes(x = group, y = obs)) + 
  geom_point() + 
  facet_wrap(~.iteration)
```


## in a hierarchical context

```{r}
ordinal_contr_hier <- cmdstan_model(here::here(
  "posts/2023-08-29-ordinal-moments/ordinal_contr_hier.stan"), pedantic = TRUE)



fake_data <- expand_grid(
  nesting(indiv_id = 1:15,
          group_id = rep(1:3, each = 5)),
  x_variable = seq(from = -15, to = 15, length.out = 12)
)

# ordinal_contr_hier$variables()$data

ord_mu_sd_sample <- ordinal_contr_hier$sample(data = list(
  n = nrow(fake_data),
  x_variable = fake_data$x_variable,
  group_id = fake_data |> 
    select(indiv_id, group_id) |> 
    distinct() |> 
    pluck("group_id"), 
  nindiv = max(fake_data$indiv_id),
  indiv_id = fake_data$indiv_id), chains = 1, refresh = 0)

some_samples <- ord_mu_sd_sample$draws(variables = "obs") |> 
  tidybayes::spread_draws(obs[i], ndraws = 12) |> 
  mutate(
    group = fake_data$group_id[i],
    x_variable = fake_data$x_variable[i]
  )

some_samples |> 
  ggplot(aes(x = x_variable, y = obs, colour = group)) + 
  geom_point() + 
  facet_wrap(~.iteration)
```


Another thing we haven't been thinking of: treatment effects on the shape parameter? 

When there is a mean-variance relationship, does this apply?
