---
title: "Probability integral transforms"
author: "Andrew"
date: 23 Oct 2022
editor: visual
---

take random numbers from a distribution

```{r}
some_numbers <- rnorm(1560, mean = 14, sd = 2.5)
hist(some_numbers)
```

Then run them through that distribution's CDF

```{r}
some_pit <- pnorm(some_numbers, mean = 14, sd = 2.5)
hist(some_pit)
```

sure enough it is uniform!

## what happens when you are wrong

let's make some curves that don't really match

```{r}
library(tidyverse)
n <- 4000
tibble(meanval = seq(from = 1, to = 14, length.out = 6),
       sd = 2.5) |> 
  expand_grid(x = seq(from = 0, to = 18, length.out = 30)) |>
  mutate(normal_dist = dnorm(x, mean = meanval, sd = sd),
         gamma_dist = dgamma(x, 
                        shape = meanval^2/sd^2,
                        rate = meanval/sd^2)) |> 
  pivot_longer(ends_with("dist"), 
               names_to = "distribution",
               values_to = "value") |> 
  ggplot(aes(x = x, y = value, colour = distribution)) + 
  geom_line()  +
  facet_wrap(~meanval)

```

We can see that the fit gets worse as the mean drops

let's simulate data from the gamma and use the PIT assuming instead it is normal:

```{r}
n <- 4000
tibble(meanval = seq(from = 1, to = 14, length.out = 6),
       sd = 2.5) |> 
  rowwise() |>
  mutate(normal_dist = list(rnorm(n, mean = meanval, sd = sd)),
         gamma_dist = list(rgamma(n, 
                        shape = meanval^2/sd^2,
                        rate = meanval/sd^2))) |> 
  pivot_longer(ends_with("dist"), 
               names_to = "distribution",
               values_to = "samples") |> 
  rowwise() |> 
  mutate(pit_samples = list(pnorm(samples, mean = meanval, sd = sd))) |> 
  select(-samples) |> 
  # filter(distribution == "gamma_dist") |> 
  unnest(pit_samples) |> 
  ggplot(aes(x = pit_samples)) + 
  geom_histogram() + 
  facet_grid(distribution~meanval)
```

let's try it with just the gamma, but changing both moments and always using the normal:

```{r}
n <- 4000
expand_grid(meanval = seq(from = 1, to = 14, length.out = 6),
       sdval = seq(from = .2, to = 7, length.out = 4)) |> 
  rowwise() |>
  mutate(gamma_dist = list(rgamma(n, 
                        shape = meanval^2/sdval^2,
                        rate = meanval/sdval^2))) |> 
  rowwise() |> 
  mutate(pit_samples = list(
    pnorm(gamma_dist,
          mean = meanval,
          sd = sdval))) |> 
  select(-gamma_dist) |> 
  # filter(distribution == "gamma_dist") |> 
  unnest(pit_samples) |> 
  ggplot(aes(x = pit_samples)) + 
  geom_histogram() + 
  facet_grid(sdval~meanval)
```


and with the lognormal

```{r}
n <- 4000
expand_grid(meanval = seq(from = 1, 
                          to = 14, 
                          length.out = 6),
       sdval = seq(from = .2, 
                   to = 7, 
                   length.out = 4)) |> 
  rowwise() |>
  mutate(
    cf = log(sdval/meanval)^2 + 1,
    lnorm_dist = list(rlnorm(n, 
                        meanlog = log(meanval) - .5*cf, 
                        sdlog = sqrt(cf))
                      )
    )|> 
  rowwise() |> 
  mutate(pit_samples = list(
    pnorm(lnorm_dist,
          mean = meanval,
          sd = sdval)
    # plnorm(lnorm_dist, 
    #        meanlog = log(meanval) - .5*cf, 
    #        sdlog = sqrt(cf))
    )) |> 
  select(-lnorm_dist) |> 
  # filter(distribution == "gamma_dist") |> 
  unnest(pit_samples) |> 
  ggplot(aes(x = pit_samples)) + 
  geom_histogram() + 
  facet_grid(sdval~meanval)
```

