---
title: "The evolution of plasticity"
author: "Andrew MacDonald and Audrey Tremblay"
description: |
  How to measure plasticity with a bayesian hierarchical model
date: 23 Nov 2022
editor: visual
categories: [UdeS, stan, brms, simulation]
---

```{r setup, eval=TRUE, message=FALSE, warning=FALSE}
options(tidyverse.quiet = TRUE)
library(targets)
library(ggplot2)
library(tidyverse)
library(tidybayes)
library(brms)
source("posts/2022-11-23-evolution-of-plasticity/evolution_of_plasticity.R")
```

## Study question

Phenotypic plasiticity is the 

## Data simulation

RARE to have more than two years per female

let's start with one female

```{r}
n <- 1
avg_csize <- 5

lifespan <- 5

general_temp <- runif(lifespan, 2, 12)

general_temp_c <- general_temp - mean(general_temp)

```

## fecundity

$$
\begin{align}
\text{eggs} &\sim \text{Poisson}(e^{\beta_0 + \beta_1*(x - \bar{x})})
\end{align}
$$

```{r}
library(tidyverse)

effet_temp <- .1

one_bird <- tibble(year = 1:lifespan,
       general_temp,
       general_temp_c,
       expected_clutch = log(avg_csize) + effet_temp * general_temp_c,
       observed_clutch = rpois(n = length(year), 
                               lambda = exp(expected_clutch)))

one_bird
```

Make a simple model to measure

```{r}
summary(glm(observed_clutch ~ general_temp_c, data = one_bird))
```

```{r}
one_bird |> 
  ggplot(aes(x = general_temp, y = observed_clutch)) + 
  geom_point()
```

## Date of laying

When do birds lay eggs?

```{r}

avg_dponte <- 138

effet_temp_dponte <- -3

one_bird_dponte <- tibble(year = 1:lifespan,
       general_temp,
       general_temp_c,
       expected_dponte = avg_dponte + effet_temp_dponte * general_temp_c,
       observed_dponte = round(rnorm(n = length(year), 
                               mean = expected_dponte,
                               sd = 5)))
```

```{r}
one_bird_dponte |> 
  ggplot(aes(x = general_temp, y= observed_dponte)) + 
  geom_point()

summary(lm(observed_dponte ~ general_temp_c, data = one_bird_dponte))
```

## combine the two

Birds which lay earlier also lay larger eggs, possibly because they are High Quality Moms.

```{r}
## population averages
pop_average_dponte <- 138
pop_average_csize <- 4

## Effect of quality
mom_quality <- 4
quality_on_dponte <- 2
quality_on_csize <- .2

```

let's observe five years for the high-quality Mom:

```{r}
quality_effects <- tibble(
  year = 1:lifespan,
  mom_quality = mom_quality,
  general_temp,
  general_temp_c,
  ## Effect of temperature -- does it depend on quality? let's say that it DOES (for now) 
  effet_temp_dponte_qual = -.7*mom_quality,
  effet_temp_csize_qual = .1*log(mom_quality),
  # csize
  mom_avg_csize = log(pop_average_csize) +  quality_on_csize*log(mom_quality),
  temp_avg_csize = exp(mom_avg_csize + effet_temp_csize_qual*general_temp_c),
  # dponte
  mom_avg_dponte = pop_average_dponte + quality_on_dponte*mom_quality,
  temp_avg_dponte = mom_avg_dponte + effet_temp_dponte_qual*general_temp_c,
  ## observations
  obs_csize = rpois(n = length(year), lambda = temp_avg_csize),
  obs_dponte = rnorm(n = length(year), mean = temp_avg_dponte, sd = 3) |> round()
)
```

Some of these values are unreasonable! we can adjust later

```{r}
#| layout-ncol: 2
#| fig-cap: 
#|   - "One female adjusting clutch size in response to temperature"
#|   - "One female adjusting laying date in repsonse to temperature"
quality_effects |> 
  ggplot(aes(x = general_temp, y = obs_csize)) + 
  geom_point()

quality_effects |> 
  ggplot(aes(x = general_temp, y = obs_dponte)) + geom_point()
```

## Multiple females

We can repeat this process for multiple females at once! let's wrap it in a function to make it easier to work with.

```{r}
simulate_many_moms
many_moms_temperature <- simulate_many_moms()
```


let's plot it!

```{r}
#| layout-ncol: 2
#| fig-cap: 
#|   - "simulations for all females in all years"
#|   - "simple linear models fit to each female, showing the variation in slopes (i.e. in plasticity"
many_moms_temperature |> 
  ggplot(aes(x = general_temp, y = obs_dponte)) + 
  geom_point()

many_moms_temperature |> 
  ggplot(aes(x = general_temp, y = obs_dponte, group = idF1)) + 
  stat_smooth(method = "lm", se = FALSE)
```

```{r}
many_moms_temperature |> 
  ggplot(aes(x = general_temp, y = obs_csize)) + 
  geom_point()

many_moms_temperature |> 
  ggplot(aes(x = general_temp, y = obs_csize, group = idF1)) + 
  stat_smooth(method = "lm", se = FALSE)
```

## model in brms


## fitness

## notes

very small sample sizes per female -- experiment with this (2 years or more)

The data are 0-truncated: only nesting females are measured!

I think it is particularly interesting that the data are already conditioned on reproduction. 
that is, females who fail to reproduce at all don't get included in the dataset. 
What effects could this have on our ability to detect interactions? 
