---
title: "Minimum observed"
author: "Andrew MacDonald"
description: |
  Sometimes we have slightly-known knowns.
date: 11 Nov 2022
editor: source
categories: [QCBS, stan]
draft: true
---

```{r setup, eval=TRUE, message=FALSE, warning=FALSE}
library(targets)
library(ggplot2)
library(tidyverse)
library(tidybayes)
library(cmdstanr)
```

## Site history effects on growth afterwards

Humans cut down forests and use that ground for something else. 


I want to imaine a scenario where you know tht total duration of the land use at a site but tnot

like you know the minimum numberf of years that the thing was abandoned, but not the maximum. at least not in every case

then thee question becomes, how many of the sites do you need to see completely to know? 

You can pass in a vector of lower bounds no problem

```{r}
lower_bound_sim <- cmdstan_model(here::here("posts/2023-09-28-interval-unknown/lower_bound_sim.stan"))

lb_samples <- lower_bound_sim$sample(data = list(n = 7, 
                                   lowers = c(-10, 2, 0, 2, 2.5, -0.6, -3)))

lb_samples$draws() |> 
  gather_rvars(alpha[i]) |> 
  ggplot(aes(y = i, dist = .value)) + 
  stat_halfeyeh()

```



## scenario

42 places to work, each one with a different history

```{r}
n <- 42
years <- 1970:2000
# random start years
startyrs <- sample(years, n, replace = TRUE)
## they are all no longer clearcut, so final year is <2000
library(purrr)
durations <- map_dbl(2001 - startyrs, sample, size = 1)

tibble(site_id = 1:n,
       start_year = startyrs,
       end_year = start_year + durations) |> 
  ggplot(aes(y = site_id, yend = site_id, x = start_year, xend = end_year)) + geom_segment() + 
  geom_vline(xintercept = 1987, col = "red", lwd = 2)

```


The idea is that each site has land-use history for various points in time
